---
title: "Introduction to birddog"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to birddog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  tidy = FALSE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
set.seed(1) # for reproducibility in stochastic steps
```

## Overview

`birddog` helps you detect emergence and trace trajectories in scientific literature and patents.
It reads datasets from OpenAlex and Web of Science (WoS), builds citation-based networks, identifies groups, and summarizes their dynamics.

A stable release is planned for CRAN. The development version is available on GitHub: <https://github.com/roneyfraga/birddog>.

## Installation

```{r eval = T}
# install.packages("devtools")
# devtools::install_github("roneyfraga/birddog")

library(birddog)
```

## Data sources

- `birddog` supports:
  - [OpenAlex](https://openalex.org/): browser search with CSV export, or API via `{openalexR}`.
  - [Web of Science](https://www.webofscience.com/wos/woscc/smart-search): multiple export formats (`.bib`, `.ris`, plain-text `.txt`, tab-delimited `.txt`).

### OpenAlex via API or CSV

You can paste a URL from openalex.org and prefix it with `https://api.` to obtain the API endpoint.

```{r eval = FALSE}

# install.packages("openalexR")
library(openalexR)

# Example: all publications in the Journal of Evolutionary Economics
url_web <- "https://openalex.org/works?page=1&filter=primary_location.source.id:s121026525"
url_api <- "https://api.openalex.org/works?page=1&filter=primary_location.source.id:s121026525"

openalexR::oa_request(query_url = url_api) |>
  openalexR::oa2df(entity = "works") |>
  birddog::read_openalex(format = "api") ->
  file

M <- birddog::read_openalex(file, format = "api")
```

### Web of Science (WoS)

WoS allows exporting in several formats. `birddog` can read:

```{r eval = FALSE}

# openalex: csv
M <- birddog::read_openalex('http://roneyfraga.com/volume/keep_it/birddog-data/openalex-works-2025-05-28T23-12-11.csv', format = "csv")

# wos: txt-plain-text
M <- birddog::read_wos('http://roneyfraga.com/volume/keep_it/birddog-data/wos-savedrecs-plain-text.txt', format = "txt-plain-text")

# wos: txt-tab-delimited
M <- birddog::read_wos('http://roneyfraga.com/volume/keep_it/birddog-data/wos-savedrecs-tab-delimited.txt', format = "txt-tab-delimited")

# wos: ris
M <- birddog::read_wos('http://roneyfraga.com/volume/keep_it/birddog-data/wos-savedrecs.ris', format = "ris")

# wos: bib
M <- birddog::read_wos('http://roneyfraga.com/volume/keep_it/birddog-data/wos-savedrecs.bib', format = "bib", normalized_names = TRUE)
```

## Example dataset

To save processing time, we’ll use a pre-saved WoS sample available in <https://roneyfraga.com/volume/keep_it/birddog-data/wos-sugarcane-m.rds>.

```
12,689 results from Web of Science Core Collection for:

"sugarcane" AND ("straw" OR "bagasse" OR "filter cake" OR "press mud" OR "pressmud cake" OR "molasses" OR "vinasse" OR "dried yeast" OR "fusel oil")
```

Download with the query above in 2023-09-27. Full query here: <https://www.webofscience.com/wos/woscc/summary/0fa06733-b4aa-4348-854d-a799cdad2c68-a711a88c/relevance/1>.

```{r eval = TRUE}

# bibs <- fs::dir_ls('~/Sync/birddog-data/bibs-sugarcane/', glob = '*.bib$')
#
# tictoc::tic()
# bibs |>
#   purrr::map(\(x) birddog::read_wos(x, format = "bib")) |>
#   dplyr::bind_rows() |>
#   dplyr::distinct(DI2, .keep_all = T) ->
#   M
# tictoc::toc()
# 62 sec

url_m <- 'https://roneyfraga.com/volume/keep_it/birddog-data/wos-sugarcane-m.rds'
M <- readRDS(url(url_m))

dplyr::glimpse(M)
```

## Build a citation network

You can build either a direct citation network or use bibliographic coupling.

Direct citation highlights time-ordered influence; bibliographic coupling captures proximity in topics via shared references.

```{r eval = TRUE}
# Direct citation
# net <- birddog::sniff_network(M, type = "direct citation")

# Bibliographic coupling
net <- birddog::sniff_network(M, type = "bibliographic coupling")

net |>
  tidygraph::activate(nodes) |>
  dplyr::select(name, AU, PY, TI, TC) |>
  dplyr::arrange(dplyr::desc(TC))
```

## Components

The analysis of components is important to eliminate disconnected documents that do not share the same bibliographic references. However, if more than one component with a high number of documents exists, it may indicate the presence of two disconnected scientific literatures.

```{r eval = TRUE}

comps <- birddog::sniff_components(net)

names(comps)

comps$components |>
  dplyr::slice_head(n = 5) |>
  gt::gt()
```

## Groups (community detection)

```{r eval = T}

birddog::sniff_groups(
  comps,
  algorithm = 'fast_greedy',
  min_group_size = 30,
  groups_short_name = TRUE) ->
  groups

names(groups)

groups$aggregate |>
  gt::gt()
```

### Group attributes

It helps to understand the structure of the groups.

```{r eval = T}

birddog::sniff_groups_attributes(
  groups,
  growth_rate_period = 2010:2022,
  show_results = FALSE) ->
  groups_attributes

names(groups_attributes)

groups_attributes$attributes_table
```

### Group content: keywords

It contributes to understanding the content of each group.

```{r eval = T}

groups_keywords <- birddog::sniff_groups_keywords(groups)

groups_keywords |>
  DT::datatable(
    rownames = FALSE,
    filter = 'bottom',
    extensions = 'Buttons',
    escape = FALSE,
    options = list(dom = 'Blfrtip', pageLength = 5)
  )
```

### Group content: NLP

This step can be time-consuming. Consider precomputing and saving results.

```{r eval = T}

# tictoc::tic()
# groups_terms <- sniff_groups_terms(groups, algorithm = 'phrase')
# tictoc::toc()
# 34 min

groups_terms <- readRDS('~/Sync/birddog-data/wos-sugarcane-groups-terms.rds')

names(groups_terms)

groups_terms$terms_table |>
  DT::datatable(
    rownames = FALSE,
    filter = 'bottom',
    extensions = 'Buttons',
    escape = FALSE,
    options = list(dom = 'Blfrtip', pageLength = 5)
  )
```

### Prestige: hubs

The calculation is slow. Be patient.

```{r eval = T}

# tictoc::tic()
# groups_hubs <- sniff_groups_hubs(groups)
# tictoc::toc()
# 19 min

groups_hubs <- readRDS('~/Sync/birddog-data/wos-sugarcane-groups-hubs.rds')

groups_hubs |>
  dplyr::filter(zone != 'noHub') |>
  dplyr::left_join(groups$network |> tidygraph::activate(nodes) |> tibble::as_tibble() |> dplyr::select(SR, PY), by = 'SR') |>
  dplyr::mutate(Zi = round(Zi, digits = 2), Pi = round(Pi, digits = 2)) |>
  dplyr::mutate(SR = paste0('<a href="https://www.webofscience.com/wos/alldb/full-record/', SR, '">', SR, '</a>')) |>
  DT::datatable(
    rownames = FALSE,
    filter = 'bottom',
    extensions = 'Buttons',
    escape = FALSE,
    options = list(dom = 'Blfrtip', pageLength = 10)
  )
```

### Group evolution (trajectories)

```{r eval = T, fig.width=10, fig.height=7}

# tictoc::tic()
# groups_cumulative <- sniff_groups_cumulative(groups)
# tictoc::toc()
# 2 min

groups_cumulative <- readRDS('~/Sync/birddog-data/wos-sugarcane-groups-cumulative.rds')

suppressMessages({
  groups_cumulative_trajectories <- birddog::sniff_groups_trajectories(groups_cumulative)
})

plot_group_trajectories_2d(
  groups_cumulative_trajectories,
  group = 'component1_g03',
  label_vertical_position = -2
)

plot_group_trajectories_3d(
  groups_cumulative_trajectories,
  group = 'component1_g03'
)
```

### Citation growth per document

```{r eval = T}

# tictoc::tic()
# groups_cumulative_citations <- sniff_groups_cumulative_citations(groups, min_citations = 2)
# tictoc::toc()
# 11 min

groups_cumulative_citations <- rio::import('~/Sync/birddog-data/wos-sugarcane-groups-cumulative-citations.rds')

groups_cumulative_citations |>
  purrr::map(\(x)
    x |>
      dplyr::select(- citations_by_year) |>
      dplyr::arrange(dplyr::desc(growth_power)) |>
      dplyr::slice_head(n = 50)) |>
  dplyr::bind_rows() |>
  dplyr::mutate(SR = paste0('<a href="https://www.webofscience.com/wos/alldb/full-record/', SR, '">', SR, '</a>')) |>
  DT::datatable(
    rownames = FALSE,
    filter = 'bottom',
    extensions = 'Buttons',
    escape = FALSE,
    options = list(dom = 'Blfrtip', pageLength = 10)
  )
```

### Topic modeling (STM)

Detect topics within a group with Structural Topic Modeling. Here, we create topics (sub-groups) based on linguistic similarities.

```{r eval = T}

# g01

# tictoc::tic()
# groups_stm_prepare_g01 <- sniff_groups_stm_prepare(groups, group_to_stm = 'g01')
# tictoc::toc()
# 21 min

groups_stm_prepare <- readRDS('~/Sync/birddog-data/wos-sugarcane-groups-stm-prepare-g01.rds')
names(groups_stm_prepare)

groups_stm_prepare$plots

# tictoc::tic()
# groups_stm_run <- sniff_groups_stm_run(groups_stm_prepare, k_topics = 18, n_top_documents = 20)
# tictoc::toc()
# 35 sec

groups_stm_run <- readRDS('~/Sync/birddog-data/wos-sugarcane-groups-stm-run-g01.rds')

groups_stm_run$topic_proportion |>
  dplyr::mutate(topic_proportion = round(topic_proportion, 3)) |>
  DT::datatable(
    caption = 'g01',
    rownames = FALSE,
    filter = 'bottom',
    extensions = 'Buttons',
    escape = FALSE,
    options = list(dom = 'Blfrtip', pageLength = 10)
  )

groups_stm_run$top_documents |>
  dplyr::left_join(M |> dplyr::select(document = DI2, SR), by = dplyr::join_by(document)) |>
  dplyr::mutate(SR = paste0('<a href="https://www.webofscience.com/wos/alldb/full-record/', SR, '">', SR, '</a>')) |>
  dplyr::select(SR, topic, gamma, title) |>
  DT::datatable(
    caption = 'g01',
    rownames = FALSE,
    filter = 'bottom',
    extensions = 'Buttons',
    escape = FALSE,
    options = list(dom = 'Blfrtip', pageLength = 10)
  )
```

## Session info

```{r eval = T}

sessioninfo::session_info()$platform |>
  unlist() |>
  as.data.frame() |>
  tibble::rownames_to_column() |>
  setNames(c("Setting", "Value")) |>
  gt::gt()
```

## Hardware

- Hostname: `r Sys.info() |> data.frame() |> janitor::clean_names() |> tibble::rownames_to_column('var') |> dplyr::filter(var == 'nodename') |> dplyr::pull(sys_info)`
- Processor: `r benchmarkme::get_cpu()[[2]]`.
- RAM: `r round(as.numeric(benchmarkme::get_ram()) / 1073741824, digits = 1)` Gigabit.
- Storage: 2 SSD's in `raid0` for data and 1 SSD for the OS.

